Forecasting the old-age dependency ratio to determine the best pension age

### 2020-09-10 PensionAge.Rmd
proselint and styler

---
title: "Forecasting the old-age dependency ratio to determine the best pension age"
author:
- familyname: Hyndman
  othernames: Rob J
  address: Department of Econometrics & Business Statistics\newline Monash University, Clayton VIC 3800, Australia
  email: Rob.Hyndman@monash.edu
  correspondingauthor: true
- familyname: Zeng
  othernames: Yijun
  address: Monash University
- familyname: Shang
  othernames: Han Lin
  address: Macquarie University
abstract: "We forecast the old-age dependency ratio for Australia under various pension age proposals, and estimate a pension age scheme that will provide a stable old-age dependency ratio at a specified level. Our approach involves a stochastic population forecasting method based on coherent functional data models for mortality, fertility and net migration, which we use to simulate future age-structures of the population. We apply these ideas to Australian historical demographic data and the results suggest that the pension age should be increased to 67 by 2021, 70 by 2032 and 75 by 2058 in order to maintain the old-age dependency ratio at the 2014 level, which is 21%. This approach can easily be extended to other target levels of the old-aged dependency ratio and other countries."
keywords: "coherent forecasts, demographic components, functional time series model, pension and insurance"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: pensionage.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    includes:
      in_header: preamble.tex
---

@@

Age pension in Australia is income support and provides access to a range of concessions for older Australians who have passed means-testing. For example, it provides single home-owners with a real income of $23,889 p.a and additional allowances to cover expenses such as rent or pharmaceuticals^[The pension amount is adjusted every 6 months for inflation]. The means test ensures that retirees with enough personal savings will become ineligible for the public provision by reducing the pension at fixed levels of pay as personal wealth increases. The age pension in Australia is very generous, even with the means test. For instance, a couple holding assets of more than one million dollars is still eligible for a partial age pension.

@@

All supportive arguments for increasing the pension age in Australia have only justified the need to increase the pension age. No justification on the pace or degree of the change has been provided. Also, the government has not disclosed the method it used to determine the new pension age. It is therefore not surprising that the adjustment of the pension age has been questioned by the public. A consistent method to determine the appropriate pension age would be more persuasive in general. For example, it is preferred by OECD countries that pension age should be linked to life expectancy [@OECD12]. There is a practical issue as to how to link life expectancy with pension ages given that there is no information on future mortality rates [@PC13]. @CEDA07 [p7] proposed directly linking pension age with life expectancy in the way that "pension age should increase by approximately 50% of any increase in life expectancy". However, using a "rule of thumb" approach might not be ideal, as 50% of the change in life expectancy may not be a good adjustment to pension age in the long term. Our study suggests an approach, linking the pension age with the aged dependency ratio, based on forecasts of future mortality and fertility rates. Our approach has the advantage that it can be used to justify the change in pension age and directly links the pension age with our ultimate goal of a stable financial burden for the public.

@@

In population forecasting, two very different approaches: a deterministic approach (sometimes called a quasi-statistical approach) and a stochastic approach (a purely statistical approach) are frequently used. For instance, the Australian Bureau of Statistics (ABS) is using a deterministic approach for its demographic projection, while some other countries or institutes forecast demographic trends based on purely statistical approaches such as Statistics New Zealand and the United States Census Bureau. In the deterministic approach, experts project future demographic changes using judgement-based scenarios. It ignores the non-demographic factors such as major pandemics or wars, although historically they are important factors that affect demographic behaviour [@PC13]. In stochastic approaches, the Lee-Carter (LC) method [@LC92] is the most prominent demographic forecasting method. It now has several variants and extensions that relax some of the assumptions of the basic Lee-Carter method [see, e.g., @SBH11]. Although statistical forecasting models are rigorous, they still leave a basis for judgment-based adaptation, as they sometimes produce unrealistic projections of demographic factors such as long-run projections of mortality rates [@PC13].

Therefore, some official agencies decide to mix the two approaches in their forecasts in order to provide a supplement to both approaches. For instance, the most recent forecast of the population in Australia by the @PC13 was based on basic Lee-Carter methods plus some judgement-based adaptations to address the characteristics of demographic trends that the Lee-Carter method failed to capture. The approach of forecasting used in this study is very similar to the @PC13. The Hyndman-Ullah method (2007) is used as a basic forecasting approach, which is a comprehensive extension of the Lee-Carter method. In addition, an adapted product-ratio method is used to eliminate the potential long-term divergence between female and male mortality rates. A more detailed review of forecasting methods is given below.

@@

Historical central death rates and start-year and mid-year populations of residents in Australia by sex and age in single years for age group 0--99 and 100+ have been sourced from the @HMD for 1921--2018. The start-year population for 2019 is also used to estimate net migration. Data for age-specific annual fertility rates by single years of age for 15--49 over the period 1950--2018 was obtained from the @ABS12. Even though annual mortality rates are available from 1921 to 2018, only data collected after 1950 is used for two reasons. First, the period between 1921--1950 has very different mortality patterns from later years because of wars and epidemics [@HB08]. Since the model for this study is valid under the assumption that extrapolative methods can be used, it seemed best to delete the less relevant data, which might influence estimations of parameters if we include them. Second, fertility rates are only available from 1950, so the mortality rate was cut to fit the same period.

@@

The first estimated basis function $\widehat{\phi}_1(x)$ measures the different effects of the trend over time across ages. It indicates the general pattern of fertility has very little effect on females of age 15. This makes intuitive sense because most early pregnancies are accidental. Therefore, any of the previously discussed events has no effect on early pregnancies. In addition, there are two spikes in the first basic function, which suggests that the historical trend has a distinct impact on females aged around their 20s and 40s. Furthermore, the blue line is the fitted value of future coefficients (i.e., $\widehat{\beta}_{n+h,1}$, for $h=1,2,\dots,50$) and the yellow shading shows their 80% pointwise prediction interval according to the fitted ARIMA model of first coefficients $\left\{\widehat{\beta}_{t,1}\right\}$. Other terms further explain variation that has not been captured by the first term, but less economic explanations can be given for them.

@@

n our sample, $n=2018-1950+1=69$ and we set $k=25$ and $H=25$ to perform the test. The mean absolute error of $h$-step-ahead forecast of OADR for $h=1,2,\dots,25$ is shown in Figure \@ref(fig:rollingtest). From this Figure, we observe that the mean absolute error is increasing as $h$ increases, which is reasonable as we have less relevant information to make a forecast for further apart OADR. Overall, the mean absolute error is very small. For example, the mean absolute error for 25-step-ahead forecast is 1.6%, which is very small if compared to the desired level of OADR, which is at 21%.

@@

```{r OADRwithoptimal, fig.cap="The mean forecast of OADR associated with target pension age scheme (listed in Table 1) shown in blue, along with the desired OADR level in red (i.e., mean forecast of OADR in 2018). The 80% prediction interval of the forecast of OADR is shown as blue shading.", dependson="OADRat65"}
## plots of OADR with plausible pension age schemes
OADR_optimal %>%
  ggplot(aes(x = Year)) +
  geom_hline(yintercept = target, col = "red") +
  geom_ribbon(aes(ymin = Lo, ymax = Hi), fill = "#7d7def") +
  geom_line(aes(y=OADR), col="#0000cc") +
  geom_line(data=OADR_history, aes(y=OADR)) +
  ggtitle("OADR forecast estimates for optimal pension ages") +
  ylab("Old age dependency ratio")
```

```{r OADRboundaries, fig.cap="The 80% prediction interval forecast of OADR (in blue) associated with the upper and lower boundary of plausible pension age schemes (listed in Table 2), along with desired OADR level (in red).", dependson="OADRat65"}
bind_rows(
    OADR_upper %>% mutate(Series="Upper boundary"),
    OADR_lower %>% mutate(Series="Lower boundary"),
  ) %>%
  ggplot(aes(x = Year)) +
  geom_hline(yintercept = target, col = "red") +
  geom_ribbon(aes(ymin = Lo, ymax = Hi), fill = "#7d7def") +
  geom_line(aes(y=OADR), col="#0000cc") +
  geom_line(data=OADR_history, aes(y=OADR)) +
  facet_grid(. ~ Series) +
  ggtitle("OADR forecasts for boundary pension age schemes") +
  ylab("Old age dependency ratio")
```

```{r comparisonofpensionages, fig.cap="The comparison between the target pension age scheme (in green) and real pension age scheme that has been set or proposed by the government, along with the 80% confidence interval of the target pension age shown in the green shade. The solid red line indicates the changes to the pension age from 65 to 67, announced in 2009.", dependson="OADRat65"}
pension.fc <- structure(
  list(mean=ts(pension_age_optimal$Age, start=min(pension_age_optimal$Year)),
       lower=ts(pension_age_optimal$Lo80, start=min(pension_age_optimal$Year)),
       upper=ts(pension_age_optimal$Hi80, start=min(pension_age_optimal$Year)),
       level=80),
  class="forecast")
ggplot() +
  autolayer(pension.fc, series = "Pension age to give OADR=0.21") +
  autolayer(
    ts(pension_age %>% filter(Policy=="Approved", Year >= 1995) %>% pull(Age),
       start=1995), series = "Approved change"
  ) +
  autolayer(
    ts(pension_age %>% filter(Policy=="Current", Year >= 1995) %>% pull(Age),
       start=1995), series = "Pension age at 65"
  )  +
  xlab("Year") + ylab("Age of retirement")
```

@@

The upper and lower boundaries of the plausible pension age schemes (also called the 80 % confidence interval of the target pension age scheme in this paper) are reported in Table 2, and their corresponding OADR simulation results are shown in Figure \@ref(fig:OADRboundaries). It can be seen that the lower (upper) limit of the 80% prediction interval of OADR associated with the lower (upper) boundary of the 80% confidence interval  pension age scheme is entirely stable around the desired level. Note the negative relationship between OADR and pension age. It suggests, on the one hand, that any pension age higher than the upper boundary in Table 2 is deemed to be undesirable because the age is too high to allow the age pension to provide enough welfare support or insurance to the aged. On the other hand, any pension age scheme lower than the lower boundary of the pension age reported in Table 2 will also be regarded as unreasonable, because it sacrifices the interests of the next generation to provide a very generous age pension immediately.

@@

```{r colophon, cache=FALSE}
options(width=100)
## Which R packages and versions?
devtools::session_info()
```

>>>>

---
title: "Forecasting the old-age dependency ratio to determine the best pension age"
author:
- familyname: Hyndman
  othernames: Rob J
  address: Department of Econometrics & Business Statistics\newline Monash University, Clayton VIC 3800, Australia
  email: Rob.Hyndman@monash.edu
  correspondingauthor: true
- familyname: Zeng
  othernames: Yijun
  address: Monash University
- familyname: Shang
  othernames: Han Lin
  address: Macquarie University
abstract: "We forecast the old-age dependency ratio for Australia under various pension age proposals, and estimate a pension age scheme that will provide a stable old-age dependency ratio at a specified level. Our approach involves a stochastic population forecasting method based on coherent functional data models for mortality, fertility and net migration, which we use to simulate future age-structures of the population. We apply these ideas to Australian historical demographic data and the results suggest that the pension age should be increased to 67 by 2021, 70 by 2032 and 75 by 2058 in order to maintain the old-age dependency ratio at the 2018 level, which is 21%. This approach can easily be extended to other target levels of the old-aged dependency ratio and other countries."
keywords: "coherent forecasts, demographic components, functional time series model, pension and insurance"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: pensionage.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    includes:
      in_header: preamble.tex
---

@@

Age pension in Australia is income support and provides access to a range of concessions for older Australians who have passed means-testing. For example, it provides single home-owners with a real income of $23,889 p.a and additional allowances to cover expenses such as rent or pharmaceuticals^[The pension amount is adjusted every 6 months for inflation]. The means test ensures that retirees with enough personal savings will become ineligible for the public provision by reducing the pension at fixed levels of pay as personal wealth increases. The age pension in Australia is generous, even with the means test. For instance, a couple holding assets of more than one million dollars is still eligible for a partial age pension.

@@

All supportive arguments for increasing the pension age in Australia have only justified the need to increase the pension age. No justification on the pace or degree of the change has been provided. Also, the government has not disclosed the method it used to determine the new pension age. It is therefore not surprising that the adjustment of the pension age has been questioned by the public. A consistent method to determine the appropriate pension age would be more persuasive in general. For example, it is preferred by OECD countries that pension age should be linked to life expectancy [@OECD12]. There is a practical issue as to how to link life expectancy with pension ages given that there is no information on future mortality rates [@PC13]. @CEDA07 [p7] proposed directly linking pension age with life expectancy in the way that "pension age should increase by approximately 50% of any increase in life expectancy". However, using a "rule of thumb" approach might not be ideal, as 50% of the change in life expectancy may not be a good adjustment to pension age in the long term. Our study suggests an approach, linking the pension age with the aged dependency ratio, based on forecasts of future mortality and fertility rates. Our approach has the advantage that it can be used to justify the change in pension age and directly links the pension age with the goal of a stable financial burden for the public.

@@

In population forecasting, two different approaches: a deterministic approach (sometimes called a quasi-statistical approach) and a stochastic approach (a purely statistical approach) are frequently used. For instance, the Australian Bureau of Statistics (ABS) is using a deterministic approach for its demographic projection, while some other countries or institutes forecast demographic trends based on purely statistical approaches such as Statistics New Zealand and the United States Census Bureau. In the deterministic approach, experts project future demographic changes using judgement-based scenarios. It ignores the non-demographic factors such as major pandemics or wars, although historically they are important factors that affect demographic behaviour [@PC13]. In stochastic approaches, the Lee-Carter (LC) method [@LC92] is the most prominent demographic forecasting method. It now has several variants and extensions that relax some of the assumptions of the basic Lee-Carter method [see, e.g., @SBH11]. Although statistical forecasting models are rigorous, they still leave a basis for judgment-based adaptation, as they sometimes produce unrealistic projections of demographic factors such as long-run projections of mortality rates [@PC13].

Therefore, some official agencies decide to mix the two approaches in their forecasts in order to provide a supplement to both approaches. For instance, the most recent forecast of the population in Australia by the @PC13 was based on basic Lee-Carter methods plus some judgement-based adaptations to address the characteristics of demographic trends that the Lee-Carter method failed to capture. The approach of forecasting used in this study is similar to the @PC13. The Hyndman-Ullah method (2007) is used as a basic forecasting approach, which is a comprehensive extension of the Lee-Carter method. In addition, an adapted product-ratio method is used to eliminate the potential long-term divergence between female and male mortality rates. A more detailed review of forecasting methods is given below.

@@

Historical central death rates and start-year and mid-year populations of residents in Australia by sex and age in single years for age group 0--99 and 100+ have been sourced from the @HMD for 1921--2018. The start-year population for 2019 is also used to estimate net migration. Data for age-specific annual fertility rates by single years of age for 15--49 over the period 1950--2018 was obtained from the @ABS12. Even though annual mortality rates are available from 1921 to 2018, only data collected after 1950 is used for two reasons. First, the period between 1921--1950 has different mortality patterns from later years because of wars and epidemics [@HB08]. Since the model for this study is valid under the assumption that extrapolative methods can be used, it seemed best to delete the less relevant data, which might influence estimations of parameters if we include them. Second, fertility rates are only available from 1950, so the mortality rate was cut to fit the same period.

@@

The first estimated basis function $\widehat{\phi}_1(x)$ measures the different effects of the trend over time across ages. It indicates the general pattern of fertility has little effect on females of age 15. This makes intuitive sense because most early pregnancies are accidental. Therefore, any of the previously discussed events has no effect on early pregnancies. In addition, there are two spikes in the first basic function, which suggests that the historical trend has a distinct impact on females aged around their 20s and 40s. Furthermore, the blue line is the fitted value of future coefficients (i.e., $\widehat{\beta}_{n+h,1}$, for $h=1,2,\dots,50$) and the yellow shading shows their 80% pointwise prediction interval according to the fitted ARIMA model of first coefficients $\left\{\widehat{\beta}_{t,1}\right\}$. Other terms further explain variation that has not been captured by the first term, but less economic explanations can be given for them.

@@

n our sample, $n=2018-1950+1=69$ and we set $k=25$ and $H=25$ to perform the test. The mean absolute error of $h$-step-ahead forecast of OADR for $h=1,2,\dots,25$ is shown in Figure \@ref(fig:rollingtest). From this Figure, we observe that the mean absolute error is increasing as $h$ increases, which is reasonable as we have less relevant information to make a forecast for further apart OADR. Overall, the mean absolute error is small. For example, the mean absolute error for 25-step-ahead forecast is 1.6%, which is small if compared to the desired level of OADR, which is at 21%.

@@

```{r OADRwithoptimal, fig.cap="The mean forecast of OADR associated with target pension age scheme (listed in Table 1) shown in blue, along with the desired OADR level in red (i.e., mean forecast of OADR in 2018). The 80% prediction interval of the forecast of OADR is shown as blue shading.", dependson="OADRat65"}
## plots of OADR with plausible pension age schemes
OADR_optimal %>%
  ggplot(aes(x = Year)) +
  geom_hline(yintercept = target, col = "red") +
  geom_ribbon(aes(ymin = Lo, ymax = Hi), fill = "#7d7def") +
  geom_line(aes(y = OADR), col = "#0000cc") +
  geom_line(data = OADR_history, aes(y = OADR)) +
  ggtitle("OADR forecast estimates for optimal pension ages") +
  ylab("Old age dependency ratio")
```

```{r OADRboundaries, fig.cap="The 80% prediction interval forecast of OADR (in blue) associated with the upper and lower boundary of plausible pension age schemes (listed in Table 2), along with desired OADR level (in red).", dependson="OADRat65"}
bind_rows(
  OADR_upper %>% mutate(Series = "Upper boundary"),
  OADR_lower %>% mutate(Series = "Lower boundary"),
) %>%
  ggplot(aes(x = Year)) +
  geom_hline(yintercept = target, col = "red") +
  geom_ribbon(aes(ymin = Lo, ymax = Hi), fill = "#7d7def") +
  geom_line(aes(y = OADR), col = "#0000cc") +
  geom_line(data = OADR_history, aes(y = OADR)) +
  facet_grid(. ~ Series) +
  ggtitle("OADR forecasts for boundary pension age schemes") +
  ylab("Old age dependency ratio")
```

```{r comparisonofpensionages, fig.cap="The comparison between the target pension age scheme (in green) and real pension age scheme that has been set or proposed by the government, along with the 80% confidence interval of the target pension age shown in the green shade. The solid red line indicates the changes to the pension age from 65 to 67, announced in 2009.", dependson="OADRat65"}
pension.fc <- structure(
  list(
    mean = ts(pension_age_optimal$Age, start = min(pension_age_optimal$Year)),
    lower = ts(pension_age_optimal$Lo80, start = min(pension_age_optimal$Year)),
    upper = ts(pension_age_optimal$Hi80, start = min(pension_age_optimal$Year)),
    level = 80
  ),
  class = "forecast"
)
ggplot() +
  autolayer(pension.fc, series = "Pension age to give OADR=0.21") +
  autolayer(
    ts(pension_age %>% filter(Policy == "Approved", Year >= 1995) %>% pull(Age),
      start = 1995
    ),
    series = "Approved change"
  ) +
  autolayer(
    ts(pension_age %>% filter(Policy == "Current", Year >= 1995) %>% pull(Age),
      start = 1995
    ),
    series = "Pension age at 65"
  ) +
  xlab("Year") +
  ylab("Age of retirement")
```

@@

The upper and lower boundaries of the plausible pension age schemes (also called the 80 % confidence interval of the target pension age scheme in this paper) are reported in Table 2, and their corresponding OADR simulation results are shown in Figure \@ref(fig:OADRboundaries). It can be seen that the lower (upper) limit of the 80% prediction interval of OADR associated with the lower (upper) boundary of the 80% confidence interval pension age scheme is entirely stable around the desired level. Note the negative relationship between OADR and pension age. It suggests, on the one hand, that any pension age higher than the upper boundary in Table 2 is deemed to be undesirable because the age is too high to allow the age pension to provide enough welfare support or insurance to the aged. On the other hand, any pension age scheme lower than the lower boundary of the pension age reported in Table 2 will also be regarded as unreasonable, because it sacrifices the interests of the next generation to provide a generous age pension immediately.

@@

```{r colophon, cache=FALSE}
options(width = 100)
## Which R packages and versions?
devtools::session_info()
```
